\documentclass[12pt, letterpaper]{article}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{geometry}
\geometry{legalpaper, portrait, margin=1in}

\lstset{columns=fullflexible, breaklines=true}
\renewcommand\qedsymbol{$\blacksquare$}
\usetikzlibrary{automata}

\title{Converting between Sequent Systems and Truth Trees}
\author{Andrew Nazareth and Andrii Bonenberger}

\begin{document}

\maketitle
\pagebreak

\newcommand \sep \Rightarrow

\section*{Sequent System Rules (Propositional Calculus)}

The sequent system we are using has the following rules of operation. $\Gamma$ and $\Delta$ are sets (multisets) and A and B are atomics.
\subsection*{Operational Rules}
\subsubsection*{Negation}
\[
\frac{\Gamma \sep \Delta, A}{\neg A, \Gamma \sep \Delta} \neg L
\qquad \qquad
\frac{A, \Gamma \sep \Delta}{\Gamma \sep \Delta, \neg A} \neg R
\]
\subsubsection*{Conjunction}
\[
\frac{A, B, \Gamma \sep \Delta}{A \land B, \Gamma \sep \Delta} \land L
\qquad \qquad
\frac{\Gamma \sep \Delta, A \qquad \Gamma \sep \Delta, B}{\Gamma \sep \Delta, A \land B} \land R
\]
\subsubsection*{Disjunction}
\[
\frac{A, \Gamma \sep \Delta \qquad B, \Gamma \sep \Delta}{A \lor B, \Gamma \sep \Delta} \lor L
\qquad \qquad
\frac{\Gamma \sep \Delta, A, B}{\Gamma \sep \Delta, A \lor B} \lor R
\]
\subsubsection*{Implication}
\[
\frac{\Gamma \sep \Delta, A \qquad B, \Pi \sep \Lambda}{A \to B, \Gamma, \Pi \sep \Delta, \Lambda} \to L
\qquad \qquad
\frac{A, \Gamma \sep \Delta, B}{\Gamma \sep \Delta, A \to B} \to R
\]
\subsection*{Structural Rules}
\subsubsection*{Weakening}
\[
\frac{\Gamma \sep \Delta}{A, \Gamma \sep \Delta} WL
\qquad \qquad
\frac{\Gamma \sep \Delta}{\Gamma \sep \Delta, A} WR
\]
\subsubsection*{Contraction}
\[
\frac{A, A, \Gamma \sep \Delta}{A, \Gamma \sep \Delta} CL
\qquad \qquad
\frac{\Gamma \sep \Delta, A, A}{\Gamma \sep \Delta, A} CR
\]
\subsubsection*{Exchange}
\[
\frac{\Gamma, A, B, \Pi \sep \Delta}{\Gamma, B, A, \Pi \sep \Delta} XL
\qquad \qquad
\frac{\Gamma \sep \Delta, A, B, \Lambda}{\Gamma \sep \Delta, B, A, \Lambda} XR
\]
\subsection*{File Layout}
Sequent derivations are represented in code as a tree of \texttt{SequentNode} objects. Each node stores its antecedent and succedent formula lists, the rule that was applied (\texttt{SequentNodeRule} enum), and pointers to parent node(s). \texttt{UnarySequentNode} handles rules with a single premise; \texttt{BinarySequentNode} handles rules with two premises (e.g., $\land R$, $\lor L$, $\to L$). The \texttt{SequentConverter} walks this structure and produces the corresponding truth tree in Willow-compatible format.

For example, the derivation:
$$\frac{\frac{\frac{A \sep A}{B, A \sep A}WL}{A, B \sep A}XL \qquad \frac{B \sep B}{A, B \sep B}WL}{A, B\sep A \land B} \land R$$

\noindent is constructed bottom-up as a \texttt{BinarySequentNode} (for $\land R$) whose left and right parents are \texttt{UnarySequentNode} chains applying the structural rules.

\pagebreak

\section*{Truth Tree Rules (Propositional Calculus)}

Truth trees systematically test the satisfiability of a set of formulas by decomposing compound statements into simpler components, branching when multiple possibilities must be explored.

\subsection*{Decomposition Rules}

A decomposable statement is any statement that is not a literal (i.e., not an atomic or the negation of an atomic). A checkmark ($\checkmark$) is placed next to a formula after it has been decomposed.

\subsubsection*{Non-Branching (Linear) Rules}

These rules add formulas to the current branch without splitting:

\[
\frac{P \land Q}{P \quad Q} \qquad \qquad
\frac{\neg(P \lor Q)}{\neg P \quad \neg Q} \qquad \qquad
\frac{\neg(P \to Q)}{P \quad \neg Q} \qquad \qquad
\frac{\neg\neg P}{P}
\]

In each case, \textit{both} resulting formulas are added to the same branch, one after the other.

\subsubsection*{Branching Rules}

These rules split the current branch into two sub-branches:

\[
\frac{P \lor Q}{P \mid Q} \qquad \qquad
\frac{P \to Q}{\neg P \mid Q} \qquad \qquad
\frac{\neg(P \land Q)}{\neg P \mid \neg Q}
\]

\subsubsection*{Biconditional Rules}

\[
\frac{P \leftrightarrow Q}{\begin{array}{c|c} P & \neg P \\ Q & \neg Q \end{array}}
\qquad \qquad
\frac{\neg(P \leftrightarrow Q)}{\begin{array}{c|c} P & \neg P \\ \neg Q & Q \end{array}}
\]

The vertical bar ($\mid$) denotes a branch point.

\subsection*{Closure Rules}

A branch is \textbf{closed} ($\times$) if a formula and its negation both appear on that branch:
\[
\frac{\varphi \qquad \neg\varphi}{\times}
\]

A branch is \textbf{open} ($\circ$) and \textbf{finished} if all decomposable statements have been decomposed and no closure condition is met.

\subsection*{Interpretation}

\begin{itemize}[nosep]
    \item If all branches close: the root formulas are jointly unsatisfiable.
    \item If any branch is open and finished: the root formulas are jointly satisfiable.
\end{itemize}

\subsection*{Applications}

\begin{itemize}[nosep]
    \item \textbf{Validity:} To test $A_1, \ldots, A_n \vdash B$, write $A_1, \ldots, A_n, \neg B$ at the root. All branches close $\Rightarrow$ valid.
    \item \textbf{Tautology:} Write $\neg\varphi$ at the root. All branches close $\Rightarrow$ tautology.
    \item \textbf{Equivalence:} $\varphi \equiv \psi$ iff both $\{\varphi, \neg\psi\}$ and $\{\neg\varphi, \psi\}$ are unsatisfiable.
\end{itemize}

\subsection*{File Layout}
The truth tree is serialized into Willow-compatible JSON by the \texttt{TreeSerializer}. Each node is an object with the following fields:

\begin{itemize}[nosep]
    \item \texttt{id}: unique integer identifier
    \item \texttt{text}: the formula string (e.g., \texttt{"A"}, \texttt{"(A $\land$ B)"}, \texttt{"$\times$"}, \texttt{"$\circ$"})
    \item \texttt{children}: array of child node IDs
    \item \texttt{decomposition}: array of node IDs that justify this node
    \item \texttt{parent}: ID of the parent node (absent for root)
    \item \texttt{antecedent}: ID of the node this was decomposed from
    \item \texttt{premise}: boolean indicating whether this is a root-level premise
\end{itemize}

The complete output is wrapped as:
\begin{verbatim}
{
  "nodes": [ ... ],
  "options": {
    "requireAtomicContradiction": true,
    "requireAllBranchesTerminated": true,
    "lockedOptions": false
  }
}
\end{verbatim}

\pagebreak

\section*{Putting Them Together: Sequent Derivations and Truth Trees}

The central insight is that \textbf{sequent calculus derivations are truth trees turned upside down}.

\subsection*{Axiom and Closure}

The LK initial sequent $A \sep A$ corresponds to truth tree closure. By $\neg L$, we derive $\neg A, A \sep$ (i.e., $\neg A, A \sep \bot$). In truth trees, both $A$ and $\neg A$ on the same branch triggers closure ($\times$).

\[
A \sep A \quad \approx \quad
\begin{array}{c}
A \\ \neg A \\ \times
\end{array}
\]

\subsection*{Structural Rules}

Structural rules in LK have no formal counterparts in truth trees, but are reflected informally:

\begin{itemize}[nosep]
    \item \textbf{Contraction} $\approx$ A formula can be reused on the same branch
    \item \textbf{Weakening} $\approx$ Redundant/unused formulas on a branch
    \item \textbf{Exchange} $\approx$ Decomposition order is irrelevant
\end{itemize}

\subsection*{Negation}

$\neg$ rules in LK are reflected in truth trees when applied twice (double negation $\neg\neg P \leadsto P$). Single applications of $\neg L / \neg R$ correspond to the formula's position on the left (assumed true) versus right (assumed false) of $\sep$.

\subsection*{Operational Rule Correspondences}

L-rules (preserving truth upward) trigger decomposition of the unsigned formula. R-rules (preserving falsity upward) trigger decomposition of the negated formula.

\subsubsection*{L-Rules $\to$ Decomposition of $A \circ B$}

\[
\frac{A, B \sep}{\underset{\land L}{A \land B \sep}}
\quad \approx \quad
\begin{array}{c} A \land B \\ A \\ B \end{array}
\qquad\qquad
\frac{A \sep \quad B \sep}{\underset{\lor L}{A \lor B \sep}}
\quad \approx \quad
\begin{array}{c} A \lor B \\ \swarrow \quad \searrow \\ A \qquad B \end{array}
\]

\[
\frac{\sep A \quad B \sep}{\underset{\to L}{A \to B \sep}}
\quad \approx \quad
\begin{array}{c} A \to B \\ \swarrow \quad \searrow \\ \neg A \qquad B \end{array}
\]

\subsubsection*{R-Rules $\to$ Decomposition of $\neg(A \circ B)$}

\[
\frac{\sep A \quad \sep B}{\underset{\land R}{\sep A \land B}}
\quad \approx \quad
\begin{array}{c} \neg(A \land B) \\ \swarrow \quad \searrow \\ \neg A \qquad \neg B \end{array}
\qquad\qquad
\frac{\sep A, B}{\underset{\lor R}{\sep A \lor B}}
\quad \approx \quad
\begin{array}{c} \neg(A \lor B) \\ \neg A \\ \neg B \end{array}
\]

\[
\frac{A \sep B}{\underset{\to R}{\sep A \to B}}
\quad \approx \quad
\begin{array}{c} \neg(A \to B) \\ A \\ \neg B \end{array}
\]

\subsection*{Summary Table}

\begin{center}
\begin{tabular}{l|l|l}
\textbf{Sequent Rule} & \textbf{Truth Tree Operation} & \textbf{Type} \\
\hline
$\land L$ & Decompose $A \land B$ (non-branching) & Linear \\
$\lor L$ & Decompose $A \lor B$ (branching) & Branch \\
$\to L$ & Decompose $A \to B$ (branching) & Branch \\
$\land R$ & Decompose $\neg(A \land B)$ (branching) & Branch \\
$\lor R$ & Decompose $\neg(A \lor B)$ (non-branching) & Linear \\
$\to R$ & Decompose $\neg(A \to B)$ (non-branching) & Linear \\
$\neg L / \neg R$ & Double negation (when applied twice) & Linear \\
Axiom $A \sep A$ & Closure ($\times$) & Leaf \\
\end{tabular}
\end{center}

\subsection*{Note on $\land L$ and $\lor R$}

For this project, we define $\land L$ and $\lor R$ as follows:
\[
\frac{A, B, \Gamma \sep \Delta}{A \land B, \Gamma \sep \Delta} \land L
\qquad \qquad
\frac{\Gamma \sep \Delta, A, B}{\Gamma \sep \Delta, A \lor B} \lor R
\]
These versions preserve both operands in the premise, making them reversible (the inference line is a biconditional). This avoids the ``lossy'' behavior of the standard formulation, where one operand is discarded when reading bottom-to-top.

\end{document}
